Discussion

### 2024-11-25 datapreptrynew.Rmd
prettyplotsindescriptives
firstfinalpushmaybe

```{r data}
scholars <- fload("./data/processed/scholars_20240924.rda") #make sure to lead fload first ofcourse
```

@@

# USE not all but first or last so the ties are directed from "perspective" of first author of the publications 
# usually first most prestige

@@

 note to self:look at publications website (sunbelt workshop)

# Gender

## saving data for each department 
This part, despite being lengthy and principally unnecessary, needed to stay in my code for the sake of transparency. AFterall, this code is what accumulated in the gender-data of each department in my data set.

```{r}
#inspecting scholars_20240924.rda object
#df_ego <- do.call(rbind.data.frame, scholars$demographics)
#unique(df_ego$discipline.22)

@@

#### soc department save

@@

#### polsci department save

@@

### gender - sociology department

@@

### gender - polsci

@@

### save gender data sets

@@

### Combine data (gender)

@@

>>>>

```{r data}
scholars <- fload("./data/processed/scholars_20240924.rda") 
```

@@

@@

# Saving data

This part, despite being lengthy and principally unnecessary, needed to stay in my code for the sake of transparency. AFterall, this code and the following code on this page is what accumulated in the gender-data of each department in my data set.

```{r}

@@

## Sociology Department

@@

## Political Science Department

@@

# Gender
### Sociology Department SCholars

@@

### Politicla Science Scholars

@@

Save gender data

@@

Combine data

@@

Throughout my code I will use two data sets called "scholars_..." and called "ego_df_...". I am aware that it would be more convenient to use a single dataset as a fondation for many analyses. However, as I went down a few rabbit holes in trying to prepare my data, a lot of my code relied on either two fo these data frames. I had passed a 'point of no return' in terms of the data frames and files I generated as re-doing this part yielded in messing up most of the code for other parts of this project. Hence, the two data sets.

### 2024-11-25 finaldata.Rmd
firstfinalpushmaybe

---
title: "Data & Analysis"
author: "Verooo"
date: "2024-11-25"
output: html_document
bibliography: references.bib
csl: apa.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

<!-- This work? -->

@@

The data for this study was obtained by collecting information from all Sociology and Political Science scholars at all Dutch institutions. During the initial phase of data collection, all important information regarding the author's position, university affiliation, names, IDs, departments, and more was gathered. The coordinator of this course, Jochem Tolsma, has provided a dataset with said information to be used in this project (**source sunbelt?** ). 'OpenAlex' was used to scrape all published works, including important work-related information such as work id, concepts, citations, abstracts, and subjects, up until 2024 (@Priem2022). From this data, I created a final dataset (n=101) that included all authors from the Sociology and Political Science departments at Radboud University within the time frame of 2015 to 2018. I also included a second wave of data, pertainning to 2019-2024, which is purely included for the sake of running the simulations needed to analyse the data, while the focus remained on the first wave (2015-2018) only. Based on this data collaboration networks were constructed to capture whether two scholars in the sample had collaborated on a publication during a given wave.

@@

SIENA, which stands for *Simulation Investigation for Empirical Network Analysis*, is a software tool designed to perform statistical estimation of models for longitudinal social network data. SIENA version 4.0, also known as RSiena, is a package for the R statistical software and was developed by Ruth Ripley and Krists Boitmanis, in collaboration with Tom Snijders (@RSiena_package, @RSiena_manual). The use of RSiena for this project is highly based on the practical information provided in "[Social Network Analysis for Social Scientists](https://jochemtolsma.github.io/SNA-4-Social-Scientists/)**"** (@SNASS).

>>>>

---
title: "Data & Analysis"
author: "Verooo"
date: "2024-11-23"
output: html_document
bibliography: references.bib
csl: apa.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

@@

The data for this study was obtained by collecting information from all Sociology and Political Science scholars at all Dutch institutions. During the initial phase of data collection, all important information regarding the author's position, university affiliation, names, IDs, departments, and more was gathered. The coordinator of this course, Jochem Tolsma, has provided a dataset with said information to be used in this project. 'OpenAlex' was used to scrape all published works, including important work-related information such as work id, concepts, citations, abstracts, and subjects, up until 2024 (@Priem2022). From this data, I created a final dataset (n=101) that included all authors from the Sociology and Political Science departments at Radboud University within the time frame of 2015 to 2018. I also included a second wave of data, pertainning to 2019-2024, which is purely included for the sake of running the simulations needed to analyse the data, while the focus remained on the first wave (2015-2018) only. Based on this data collaboration networks were constructed to capture whether two scholars in the sample had collaborated on a publication during a given wave.

@@

SIENA, which stands for *Simulation Investigation for Empirical Network Analysis*, is a software tool designed to perform statistical estimation of models for longitudinal social network data. SIENA version 4.0, also known as RSiena, is a package for the R statistical software and was developed by Ruth Ripley and Krists Boitmanis, in collaboration with Tom Snijders (@RSiena_package, @RSiena_manual). The use of RSiena for this project is highly based on the practical information provided in "[Social Network Analysis for Social Scientists](https://jochemtolsma.github.io/SNA-4-Social-Scientists/)" (@SNASS).

### 2024-11-25 finaldataprep.Rmd
firstfinalpushmaybe

---
title: "Data preparation"
author: "Verooo"
date: "2024-11-23"
output: html_document
bibliography: references.bib
csl: apa.csl
---
```{=html}
<style>
body {
text-align: left;
line-height: 1.5;
}

h1, .h1, h2, .h2, h3, .h3 {
  margin-top: 24px;
  line-height: 1;
}

</style>
```

@@

>>>>

---
title: "Data preparation"
author: "Verooo"
date: "2024-11-23"
output: html_document
bibliography: references.bib
csl: apa.csl
---

<html lang="en">

<head>

<meta charset="UTF-8">

<meta name="viewport" content="width=device-width, initial-scale=1.0">

<title>My Models</title>

<link rel="stylesheet" href="styles.css"> <!-- Link to the CSS file I created for dark mode code chunks -->

</head>

</html>

@@

# Getting started

```{r clean}
rm(list = ls())
```

## Packages

```{r package}
require(tidyverse)
require(RSiena)
require(RsienaTwoStep)

#install.packages("data.table")
require(data.table)  # mainly for faster data handling
# install.packages('xml2')
require(xml2)
# install.packages('rvest')
require(rvest)
require(igraph)

require(ggraph)
#install.packages('visNetwork')
require(visNetwork)
#install.packages('threejs')
require(threejs)
#install.packages('networkD3')
require(networkD3)

## Functions

```{r functions 1}
fpackage.check <- function(packages) {
    lapply(packages, FUN = function(x) {
        if (!require(x, character.only = TRUE)) {
            install.packages(x, dependencies = TRUE)
            library(x, character.only = TRUE)
        }
    })
}

fsave <- function(x, file = NULL, location = "./data/processed/") {
    ifelse(!dir.exists("data"), dir.create("data"), FALSE)
    ifelse(!dir.exists("data/processed"), dir.create("data/processed"), FALSE)
    if (is.null(file))
        file = deparse(substitute(x))
    datename <- substr(gsub("[:-]", "", Sys.time()), 1, 8)
    totalname <- paste(location, file, "_", datename, ".rda", sep = "")
    save(x, file = totalname)  #need to fix if file is reloaded as input name, not as x. 
}

fload <- function(filename) {
    load(filename)
    get(ls()[ls() != "filename"])
}

fshowdf <- function(x, ...) {
    knitr::kable(x, digits = 2, "html", ...) %>%
        kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
        kableExtra::scroll_box(width = "100%", height = "300px")
}

# this is the most important one. We created it in the previous script

f_pubnets <- function(df_scholars = df, list_publications = publications, discip = "sociology"+"political science", affiliation = "RU",
    waves = list(wave1 = c(2018, 2019, 2020), wave2 = c(2021, 2022, 2023))) {

    publications <- list_publications %>%
        bind_rows() %>%
        distinct(title, .keep_all = TRUE)

    df_scholars %>%
        filter(affil1 == affiliation | affil2 == affiliation) %>%
        filter(discipline == discip) -> df_sel

    networklist <- list()
    for (wave in 1:length(waves)) {
        networklist[[wave]] <- matrix(0, nrow = nrow(df_sel), ncol = nrow(df_sel))
    }

    publicationlist <- list()
    for (wave in 1:length(waves)) {
        publicationlist[[wave]] <- publications %>%
            filter(gs_id %in% df_sel$gs_id) %>%
            filter(year %in% waves[[wave]]) %>%
            select(author) %>%
            lapply(str_split, pattern = ",")
    }

    publicationlist2 <- list()
    for (wave in 1:length(waves)) {
        publicationlist2[[wave]] <- publicationlist[[wave]]$author %>%
            # lowercase
        lapply(tolower) %>%
            # Removing diacritics
        lapply(stri_trans_general, id = "latin-ascii") %>%
            # only last name
        lapply(word, start = -1, sep = " ") %>%
            # only last last name
        lapply(word, start = -1, sep = "-")
    }

    for (wave in 1:length(waves)) {
        # let us remove all publications with only one author
        remove <- which(sapply(publicationlist2[[wave]], FUN = function(x) length(x) == 1) == TRUE)
        publicationlist2[[wave]] <- publicationlist2[[wave]][-remove]
    }

    for (wave in 1:length(waves)) {
        pubs <- publicationlist2[[wave]]
        for (ego in 1:nrow(df_sel)) {
            # which ego?
            lastname_ego <- df_sel$lastname[ego]
            # for all publications
            for (pub in 1:length(pubs)) {
                # only continue if ego is author of pub
                if (lastname_ego %in% pubs[[pub]]) {
                  aut_pot <- which.max(pubs[[pub]] %in% lastname_ego)
                  # only continue if ego is first author of pub
                  if (aut_pot == 1) {
                    # check all alters/co-authors
                    for (alter in 1:nrow(df_sel)) {
                      # which alter
                      lastname_alter <- df_sel$lastname[alter]
                      if (lastname_alter %in% pubs[[pub]]) {
                        networklist[[wave]][ego, alter] <- networklist[[wave]][ego, alter] + 1
                      }
                    }
                  }
                }
            }
        }
    }
    return(list(df = df_sel, network = networklist))
}

## More functions

```{r functions 2}
fcolnet <- function(data = scholars, university = "RU", discipline = "sociology"+"political science", waves = list(c(2015,
    2018), c(2019, 2023)), type = c("first")) {

    # step 1
    demographics <- do.call(rbind.data.frame, data$demographics)
    demographics <- demographics %>%
        mutate(Universiteit1.22 = replace(Universiteit1.22, is.na(Universiteit1.22), ""), Universiteit2.22 = replace(Universiteit2.22,
            is.na(Universiteit2.22), ""), Universiteit1.24 = replace(Universiteit1.24, is.na(Universiteit1.24),
            ""), Universiteit2.24 = replace(Universiteit2.24, is.na(Universiteit2.24), ""), discipline.22 = replace(discipline.22,
            is.na(discipline.22), ""), discipline.24 = replace(discipline.24, is.na(discipline.24), ""))

    sample <- which((demographics$Universiteit1.22 %in% university | demographics$Universiteit2.22 %in%
        university | demographics$Universiteit1.24 %in% university | demographics$Universiteit2.24 %in%
        university) & (demographics$discipline.22 %in% discipline | demographics$discipline.24 %in% discipline))

    demographics_soc <- demographics[sample, ]
    scholars_sel <- lapply(scholars, "[", sample)

    # step 2
    ids <- demographics_soc$au_id
    nwaves <- length(waves)
    nets <- array(0, dim = c(nwaves, length(ids), length(ids)), dimnames = list(wave = 1:nwaves, ids,
        ids))
    dimnames(nets)

    # step 3
    df_works <- tibble(works_id = unlist(lapply(scholars_sel$work, function(l) l$id)), works_author = unlist(lapply(scholars_sel$work,
        function(l) l$author), recursive = FALSE), works_year = unlist(lapply(scholars_sel$work, function(l) l$publication_year),
        recursive = FALSE))

    df_works <- df_works[!duplicated(df_works), ]

    # step 4
    if (type == "first") {
        for (j in 1:nwaves) {
            df_works_w <- df_works[df_works$works_year >= waves[[j]][1] & df_works$works_year <= waves[[j]][2],
                ]
            for (i in 1:nrow(df_works_w)) {
                ego <- df_works_w$works_author[i][[1]]$au_id[1]
                alters <- df_works_w$works_author[i][[1]]$au_id[-1]
                if (sum(ids %in% ego) > 0 & sum(ids %in% alters) > 0) {
                  nets[j, which(ids %in% ego), which(ids %in% alters)] <- 1
                }
            }
        }
    }

    if (type == "last") {
        for (j in 1:nwaves) {
            df_works_w <- df_works[df_works$works_year >= waves[[j]][1] & df_works$works_year <= waves[[j]][2],
                ]
            for (i in 1:nrow(df_works_w)) {
                ego <- rev(df_works_w$works_author[i][[1]]$au_id)[1]
                alters <- rev(df_works_w$works_author[i][[1]]$au_id)[-1]
                if (sum(ids %in% ego) > 0 & sum(ids %in% alters) > 0) {
                  nets[j, which(ids %in% ego), which(ids %in% alters)] <- 1
                }
            }
        }
    }

    if (type == "all") {
        for (j in 1:nwaves) {
            df_works_w <- df_works[df_works$works_year >= waves[[j]][1] & df_works$works_year <= waves[[j]][2],
                ]
            for (i in 1:nrow(df_works_w)) {
                egos <- df_works_w$works_author[i][[1]]$au_id
                if (sum(ids %in% egos) > 0) {
                  nets[j, which(ids %in% egos), which(ids %in% egos)] <- 1
                }
            }
        }
    }
    output <- list()
    output$data <- scholars_sel
    output$nets <- nets
    return(output)
}
```

## Load data

```{r data}
scholars <- fload("./data/processed/scholars_20240924.rda") 
```

```{r saving output}
# save the output of your function

test  <- fcolnet(data = scholars, 
                university = "RU", 
                discipline = c("sociology", "political science"),
                waves = list(c(2015, 2018), c(2019, 2023)), 
                type = c("all")) # all, first, last # which ties in there

# Saving data

This part, despite being lengthy and principally unnecessary, needed to stay in my code for the sake of transparency. AFterall, this code and the following code on this page is what accumulated in the gender-data of each department in my data set.

```{r}

# save the output of your function
testsoc  <- fcolnet(data = scholars, 
                university = "RU", 
                discipline = "sociology", 
                waves = list(c(2015, 2018), c(2019, 2023)),

                            type = c("first"))

testpolsci <- fcolnet(data = scholars, 
                university = "RU", 
                discipline = "political science", 
                waves = list(c(2015, 2018), c(2019, 2023)), 
                type = c("first"))
```

## Sociology Department

```{r}
soc_nets <- testsoc$nets

df_ego_demo <- testsoc$data$demographics
df_ego_oa <- testsoc$data$scholars_oa
df_ego_works <- testsoc$data$works

# DEMO
  str(df_ego_demo, 1)

  df_ego_demo <- do.call(rbind.data.frame, df_ego_demo)
  df_ego_demo[[1]]

# OA
  #str(df_ego_oa, 1)
  df_ego_oa[[1]]
  df_ego_oa <- do.call(rbind.data.frame, df_ego_oa)

  # --> EGO DATAFRAME with demo and oa (sociology)

  df_soc_ego <- left_join(df_ego_demo, df_ego_oa, by = join_by(au_id == id))

  rownames(soc_nets[1,,]) == df_soc_ego$au_id # checking if the author id's are the same as in df_ego in nets dataframe 
  # --> yes, the correct ones!!!

  df_ego_works[[1]]$author 
#authors of all the works that first scholar (here: Ana Macanovic) is part of

## Political Science Department

```{r}
# dependent
# pol_net <- sienaDependent(net_pol_array)

# dependent 
pol_nets <- testpolsci$nets

df_pol_ego_demo <- testpolsci$data$demographics
df_pol_ego_oa <- testpolsci$data$scholars_oa
df_pol_ego_works <- testpolsci$data$works

# DEMO
  #str(df_pol_ego_demo, 1)

  df_pol_ego_demo <- do.call(rbind.data.frame, df_pol_ego_demo)
  df_pol_ego_demo[[1]]

# OA
  #str(df_pol_ego_oa, 1)
  df_pol_ego_oa[[1]]
  df_pol_ego_oa <- do.call(rbind.data.frame, df_pol_ego_oa)

# --> EGO DATAFRAME with demo and oa (political science)
  df_pol_ego <- left_join(df_pol_ego_demo, df_pol_ego_oa, by = join_by(au_id == id))

  rownames(pol_nets[1,,]) == df_pol_ego$au_id # checking if the author id's are the same as in df_ego in nets dataframe 
  # --> yes, the correct ones!!!
```

# Gender
### Sociology Department SCholars

```{r gender s}
df_soc_ego$Naam # list of all names in sociology department

df_soc_ego$gender <- c(0,1,0,1,0,0,1,1,0,0,
                       0,1,1,0,1,0,1,1,1,1,
                       0,1,1,0,1,1,0,1,0,1,
                       0,1,0,0,0,0,0,0,1,0,
                       0,1,1,1,0,1,0,1,0,0) 
# manually filled in gender 0 = woman, 1 = man, based on majority given in NL voornamen bank

# check if it went smoothly 
# view(df_soc_ego[c('Naam', 'gender', 'au_id')])

# fsave(df_soc_ego)

### Politicla Science Scholars

```{r gender p}
df_pol_ego$Naam # list of all names

df_pol_ego$gender <- c(1,0,1,0,0,0,1,0,1,0,
                       1,1,0,0,1,1,1,0,0,1,
                       0,1,1,0,0,1,1,1,0,0,
                       0,0,0,1,1,0,0,1,1,0,
                       0,0,0,1,0,1,1,1,0,0,0) 
    # manually filled in gender 0 = woman, 1 = man, based on majority given in NL voornamen bank, Guangyu was not in name bank but based on the employee-picture of RU, this person presents feminine, thus I assumed 0.

# check if it went smoothly 
# view(df_pol_ego[c('Naam', 'gender', 'au_id')])

#covariate 
# gender <- coCovar(df_ego$gender)

# fsave(df_pol_ego)
```

Save gender data

```{r gender data}
soc_gender <- fload("./data/processed/df_soc_ego_20241105.rda")
pol_gender <- fload("./data/processed/df_pol_ego_20241105.rda")
```

Combine data

```{r combine gender set}
df <- test$data # make dataset with all data 
df_ego <- do.call(rbind.data.frame, df$demographics) # data set with demographic information

soc_gender <- soc_gender[,c("Naam", "gender")] # selecting only the content we need
pol_gender <- pol_gender[,c("Naam", "gender")] # selecting only the content we need

gender <- rbind(soc_gender, pol_gender) # combining the gender data sets

df_ego <- df_ego %>% 
            left_join(gender) #dcombining data set of gender with the one with demographic information 
```

# H-index

Credits to Lucan Bovens for the code.

```{r h-index}
# Add h-index and i10-index columns to df_ego
df_ego <- df_ego %>%
  mutate(h_index = NA_real_, i10_index = NA_real_)

# Get H-index and i10-index 
get_openalex_metrics <- function(openalex_id) {
  url <- paste0("https://api.openalex.org/authors/", openalex_id)
  author_data <- tryCatch({
    jsonlite::fromJSON(url)
  }, error = function(e) {
    message(paste("Error fetching data for OpenAlex ID:", openalex_id))
    return(NULL)
  })

  if (!is.null(author_data)) {
    h_index <- author_data$summary_stats$h_index
    i10_index <- author_data$summary_stats$i10_index
    works_count <- author_data$works_count
    cited_by_count <- author_data$cited_by_count
    return(list(h_index = h_index, i10_index = i10_index, works_count = works_count, cited_by_count = cited_by_count))
  } else {
    return(list(h_index = NA, i10_index = NA, works_count = NA, cited_by_count = NA))
  }
}

# Loop for each scholar 
for (i in 1:nrow(df_ego)) {
  openalex_id <- df_ego$au_id[i]  # assuming au_id is the OpenAlex ID in df_ego
  metrics <- get_openalex_metrics(openalex_id)
  df_ego$h_index[i] <- metrics$h_index
  df_ego$i10_index[i] <- metrics$i10_index
  df_ego$works_count[i] <- metrics$works_count
  df_ego$cited_by_count[i] <- metrics$cited_by_count
}

# View updated df_ego with h-index and i10-index
view(df_ego[c('au_id','Naam','gender','h_index','i10_index')])

# fsave(df_ego)

Throughout my code I will use two data sets called "scholars_..." and called "ego_df_...". I am aware that it would be more convenient to use a single dataset as a fondation for many analyses. However, as I went down a few rabbit holes in trying to prepare my data, a lot of my code relied on either two fo these data frames. I had passed a 'point of no return' in terms of the data frames and files I generated as re-doing this part yielded in messing up most of the code for other parts of this project. Hence, the two data sets.

### 2024-11-25 finaldiscussion.Rmd
firstfinalpushmaybe

The results consistently support Hypothesis 1, showing significant evidence of gender homophily across all models. The parameter for same-gender collaboration (sameX) remains positive and significant, indicating that scholars are more likely to collaborate with same-gender peers than with different-gender peers. This aligns with previous research suggesting that gendered barriers in academia, whether structural or cultural, can lead to gender clustering in collaborative networks (@Horta2022; @Kwiek2021; @McPherson2001).

Interestingly, the significance of gender clustering persists even after controlling for other covariates, such as department affiliation and h-index, demonstrating that it is a robust phenomenon in these networks. While this result is not unexpected, it emphasizes the enduring influence of gender dynamics on scholarly collaborations.

@@

Hypothesi 2 proposed that scholars in the sociology department exhibit weaker gender homophily than those in the political science department. While this hypothesis was not explicitly tested using an interaction term, comparisons across models suggest no substantial department-level differences in gender clustering. The same-gender collaboration parameter (sameX for gender) is nearly identical in models 2 and 3, indicating that the inclusion of department as a covariate does not significantly alter the pattern of same-gender collaborations.

However, the department-related parameter for ego’s tendency to initiate collaborations (egoX for department) becomes significant in model 4, revealing a lower likelihood for political scientists to initiate collaborations compared to sociologists. This could imply department-level differences in collaboration norms or opportunities, which might indirectly influence gender homophily. An interaction term between department and gender would have been useful to further investigate these nuances.

After all, hypothesis 2 was not supported. While scholars show a significant preference for collaborating within their own department, with sociologists more likely to initiate collaborations than political scientists, *no* substantial differences in gender homophily were observed in relation to scholars' departments.

@@

In terms of structural network effects, reciprocity emerges as a consistently significant factor, indicating that collaborations are more likely to be mutual. This aligns with existing theories of dyadic reciprocity in networks, suggesting that once a collaboration tie is initiated, it is likely to be reciprocated (@Snijders2010). Interestingly, the outdegree parameter (baseline likelihood of initiating a tie) fluctuates in significance across models but remains consistently negative when significant, indicating that scholars generally show a low tendency to initiate collaborations.

In line with Hypothesis 3, Gender homophily (sameX) remains significant even when controlling for these structural effects, reinforcing its importance. This finding implies that scholars' same-gender preferences in creating ties are not merely a byproduct of network structure but reflect deeper social or cultural dynamics.

@@

The results provide little to no support for Hypothesis 4, which proposed that gender homophily is less influential when potential collaborators have higher academic prestige (as measured by the h-index). The inclusion of h-index as a covariate in model 4 reveals that scholars with higher h-index scores are more likely to attract collaborations (significant positive alter effect, altX) but less likely to initiate them (significant negative ego effect, egoX). This is consistent with theories of preferential attachment and upward mobility, where prestigious scholars act as hubs, receiving collaboration requests rather than initiating them (@Broido2019; @Knobloch2013).

However, since the the same-gender parameter (sameX) remained significant and largely unchanged, the influence of genderhomophily on scholars' decisions to collaborate is not substantially impacted by scholars' h-index. This finding suggests that while prestige plays a role in shaping collaboration networks, it does mitigate the tendency of scholars to prefer same-gender collaborations over different-gendered ones. Thus, while prestigious scholars are more sought after, collaboration preferences in scholars' networks remain influenced by gender-preferences.

@@

The findings of this study highlight the interplay between structural, departmental, and individual factors in shaping co-authorship networks. Gender homophily appears to be a robust and enduring feature of these networks, resisting the influence of prestige and departmental context. This raises important questions about the structural barriers faced by scholars, particularly women, in accessing diverse collaboration opportunities.

From a policy perspective, these results underscore the need for initiatives aimed at reducing gendered collaboration clustering. For example, mentorship programs or institutional incentives that encourage cross-gender and cross-department collaborations could help lessen these pattern. Additionally, the finding that prestigious scholars are less likely to initiate collaborations suggests that researchers early int heir career, particularly those from underrepresented groups, may face challenges in accessing these high-prestige hubs. Institutions should consider strategies to facilitate connections between early-career and high-prestige scholars to promote equity in collaboration opportunities.

@@

This study has several limitations that warrant consideration. First, the lack of interaction terms between gender and department or h-index limits the ability to make strong conclusions about how these factors jointly influence collaboration patterns. Future studies should include such interactions to better understand these dynamics. Second, while the h-index serves as a useful proxy for academic prestige, this study did not control for author characteristics such as age, which may account for some of the effects found. Besides, other measures of prestige or influence, such as grant funding or editorial roles, might yield additional insights. Finally, the examined sample is arguably small, and the generalizability of these findings is further limited to the specific departments and network structures examined here.

Future research should explore longitudinal changes in collaboration patterns to assess whether gender homophily weakens over time, particularly as scholars gain prestige. Additionally, examining the role of institutional interventions, such as funding programs or diversity initiatives, could provide valuable insights into strategies for fostering more equitable collaboration networks.

>>>>

The results consistently support Hypothesis 1, showing significant evidence of gender homophily across all models. The parameter for same-gender collaboration (sameX) remains positive and significant, indicating that scholars are more likely to collaborate with same-gender peers than with different-gender peers. This aligns with previous research suggesting that gendered barriers in academia, whether structural or cultural, can lead to gender clustering in collaborative networks (@Horta2022; @Kwiek2021; @McPherson2001).\
Interestingly, the significance of gender clustering persists even after controlling for other covariates, such as department affiliation and h-index, demonstrating that it is a robust phenomenon in these networks. While this result is not unexpected, it emphasizes the enduring influence of gender dynamics on scholarly collaborations.

@@

Hypothesi 2 proposed that scholars in the sociology department exhibit weaker gender homophily than those in the political science department. While this hypothesis was not explicitly tested using an interaction term, comparisons across models suggest no substantial department-level differences in gender clustering. The same-gender collaboration parameter (sameX for gender) is nearly identical in models 2 and 3, indicating that the inclusion of department as a covariate does not significantly alter the pattern of same-gender collaborations.\
However, the department-related parameter for ego’s tendency to initiate collaborations (egoX for department) becomes significant in model 4, revealing a lower likelihood for political scientists to initiate collaborations compared to sociologists. This could imply department-level differences in collaboration norms or opportunities, which might indirectly influence gender homophily. An interaction term between department and gender would have been useful to further investigate these nuances. After all, hypothesis 2 was not supported. While scholars show a significant preference for collaborating within their own department, with sociologists more likely to initiate collaborations than political scientists, *no* substantial differences in gender homophily were observed in relation to scholars' departments.

@@

In terms of structural network effects, reciprocity emerges as a consistently significant factor, indicating that collaborations are more likely to be mutual. This aligns with existing theories of dyadic reciprocity in networks, suggesting that once a collaboration tie is initiated, it is likely to be reciprocated (@Snijders2010). Interestingly, the outdegree parameter (baseline likelihood of initiating a tie) fluctuates in significance across models but remains consistently negative when significant, indicating that scholars generally show a low tendency to initiate collaborations. In line with Hypothesis 3, Gender homophily (sameX) remains significant even when controlling for these structural effects, reinforcing its importance. This finding implies that scholars' same-gender preferences in creating ties are not merely a byproduct of network structure but reflect deeper social or cultural dynamics.

@@

The results provide little to no support for Hypothesis 4, which proposed that gender homophily is less influential when potential collaborators have higher academic prestige (as measured by the h-index). The inclusion of h-index as a covariate in model 4 reveals that scholars with higher h-index scores are more likely to attract collaborations (significant positive alter effect, altX) but less likely to initiate them (significant negative ego effect, egoX). This is consistent with theories of preferential attachment and upward mobility, where prestigious scholars act as hubs, receiving collaboration requests rather than initiating them (@Broido2019; @Knobloch2013).\
However, since the the same-gender parameter (sameX) remained significant and largely unchanged, the influence of genderhomophily on scholars' decisions to collaborate is not substantially impacted by scholars' h-index. This finding suggests that while prestige plays a role in shaping collaboration networks, it does mitigate the tendency of scholars to prefer same-gender collaborations over different-gendered ones. Thus, while prestigious scholars are more sought after, collaboration preferences in scholars' networks remain influenced by gender-preferences.

@@

The findings of this study highlight the interplay between structural, departmental, and individual factors in shaping co-authorship networks. Gender homophily appears to be a robust and enduring feature of these networks, resisting the influence of prestige and departmental context. This raises important questions about the structural barriers faced by scholars, particularly women, in accessing diverse collaboration opportunities.\
From a policy perspective, these results underscore the need for initiatives aimed at reducing gendered collaboration clustering. For example, mentorship programs or institutional incentives that encourage cross-gender and cross-department collaborations could help lessen these pattern. Additionally, the finding that prestigious scholars are less likely to initiate collaborations suggests that researchers early int heir career, particularly those from underrepresented groups, may face challenges in accessing these high-prestige hubs. Institutions should consider strategies to facilitate connections between early-career and high-prestige scholars to promote equity in collaboration opportunities.

@@

This study has several limitations that warrant consideration. First, the lack of interaction terms between gender and department or h-index limits the ability to make strong conclusions about how these factors jointly influence collaboration patterns. Future studies should include such interactions to better understand these dynamics. Second, while the h-index serves as a useful proxy for academic prestige, this study did not control for author characteristics such as age, which may account for some of the effects found. Besides, other measures of prestige or influence, such as grant funding or editorial roles, might yield additional insights. Finally, the examined sample is arguably small, and the generalizability of these findings is further limited to the specific departments and network structures examined here. \
Future research should explore longitudinal changes in collaboration patterns to assess whether gender homophily weakens over time, particularly as scholars gain prestige. Additionally, examining the role of institutional interventions, such as funding programs or diversity initiatives, could provide valuable insights into strategies for fostering more equitable collaboration networks.
